---
title: "Crop Yield"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Impact of Cover Crops on Wheat Crop Yields

Introduction

As the global population continues to grow, the demand for sustainable agricultural practices that can support increased food production is becoming more critical. 
Cover crops, a type of sustainable farming practice, are increasingly being used to improve soil health, reduce erosion, and manage weeds and pests. However, the impact of cover crops on the yields of staple crops such as wheat remains a subject of debate among farmers and researchers.
This project aims to use causal inference techniques to evaluate whether the adoption of cover crops leads to an increase in wheat yields. 
By applying methods such as inverse probability of treatment weighting (IPTW) and propensity score matching, we can better understand the true effect of cover crops while accounting for potential confounders like farm size, soil quality, and regional agricultural practices. 
The findings from this analysis will provide insights that can help guide policy and farming decisions toward more effective and sustainable agricultural practices.


# Import libraries
```{r, message=FALSE}
library(palmerpenguins)
library(tidyverse)
library(cobalt)
library(WeightIt)
library(lmtest)
library(sandwich)
library(ggplot2)
```

# Data Exploration

```{r}
# Load file as dataframe
farm_df <- read.csv("farms.csv")
```

```{r}
# Inspect dataframe
head(farm_df)
```

# Initial data observations

Outcome:

    total_yield: represents the average total yield of wheat in bushels per acre.

Treatment:

    cover_10: indicates whether at least 10% of farms in a county employ cover crops.

Predictors:

    region: geographic region where the county is located.

    total_avg: average total size of farms (thousands of acres).

    age_avg: average age of the primary farm operator (years).

    experience_avg: average number of years of farming experience of the primary farm operator (years).

    insurance_avg: average area of land with crop insurance (thousands of acres).

    easement_p: indicates the average percent of land under an easement in the county.

    conservation_till_avg: average area of land that uses conservation tillage methods (acres).

    fertilizer_per_area: average cost of fertilizers used per acre (hundreds of thousands of dollars).

# Examine Initial Overlap and Balance

Assess overlap and balance visually for a couple of variables to compare counties with AT LEAST 10% of farms using cover crops and counties with LESS THAN 10% of farms using cover crops.

I will use a balance plot for the age_avg variable. 

Do the treatment and control distributions appear to be centered in the same location and have similar spreads?

```{r}
# Balance plot for the average age
bal.plot(
  x = cover_10 ~ age_avg, #formula
  data = farm_df, #dataset
  var.name = "age_avg", #variable
  colors = c("#E69F00", "#009E73") #set fill colors
)
```

Are the proportions of counties in the treatment versus control groups similar across the four regions?

```{r}
# Balance plot for region
bal.plot(
  x = cover_10 ~ region, #formula
  data = farm_df, #dataset
  var.name = "region", #variable
  colors = c("#E69F00", "#009E73") #set fill colors
)
```

Examine the standardized mean differences (SMD) and variance ratios for all the predictor variables according to the treatment group. Check whether the balance measurements fall outside of the guidelines of ±0.1 for SMDs and between 0.5 and 2 for variance ratios.

```{r}
# Balance table to show standardized mean differences (SMD) and variance ratios
bal.tab(
  x = cover_10 ~ age_avg + region, data = farm_df,
  binary = "std", disp.v.ratio = TRUE
)
```

# Perform IPTW Procedure and Re-check Balance

Perform the inverse probability of treatment weighting (IPTW) procedure to see if we can minimize imbalance between the treatment groups before estimating the causal treatment effect."

"Next, perform an IPTW procedure using region, average farm size, average number of farms with crop insurance, and average cost of fertilizer per acre as predictors in the propensity score model. Calculate weights in order to estimate the average treatment effect (ATE).


```{r}
farm_iptw <- weightit(
  formula = cover_10 ~ region +
    total_avg + insurance_avg + fertilizer_per_area,
  data = farm_df, method = "ps", estimand = "ATE"
)

farm_iptw
```

Create a Love plot showing the SMD between treatment groups for each variable in the propensity score model. Display the SMD for binary variables and the SMD threshold of ± 0.1.

```{r}
# Love plot showing the SMD
# between treatment groups for each variable in the propensity score model
love.plot(
  farm_iptw, binary = "std", thresholds = c(m = 0.1)
)
```

# Refine Propensity Score Model

Can I improve the balance by tuning the propensity score model?

Re-run the IPTW procedure with a new propensity score model. Achieve this by removing the fertilizer_per_area variable from the model. Then add varibles:

- the average age of the primary producer
- the average experience of the primary producer
- the average percent of land under easement
- the average number of farms using conservation tillage

```{r}
farm_iptw2 <- farm_iptw2 <- weightit(
  formula = cover_10 ~ region + total_avg
  + age_avg + experience_avg
  + easement_p + conservation_till_avg
  + insurance_avg,
  data = farm_df, method = "ps", estimand = "ATE"
)

farm_iptw2
```

```{r}
# Love plot showing the SMD between
# treatment groups for each variable
# in the propensity score model
love.plot(
  farm_iptw2, binary = "std", thresholds = c(m = 0.1)
)
```

The Love plot shows SMD values closer to zero than the previous model.

Next, inspect the distribution of propensity scores between treatment groups to measue IPTW procedure performance.

```{r}
# Generate a balance plot to display
# the distribution of propensity scores in each treatment group before
# AND after the weighting process
bal.plot(
  x = farm_iptw, var.name = "prop.score", which = "both"
)

```

```{r}
bal.plot(
  x = farm_iptw2, var.name = "prop.score", which = "both"
)
```

Do the distributions of the weighted propensity scores look closer to identical, and are they overlapping each other?

# Estimate Causal Treatment Effect

Last step of the analysis: 

- Fit the final outcome model to estimate the causal effect of cover crop usage on crop yields.

1. Fit the outcome regression model with total crop yield as the outcome, cover crop usage as the treatment variable, and the other variables from the second propensity score model as the additional predictors. 
- Include the weights from the IPTW procedure in the regression model.

```{r}
yield_mod <- glm(
  total_yield ~ cover_10 + region
  + total_avg + age_avg
  + experience_avg + easement_p
  + conservation_till_avg + insurance_avg,
  data = farm_df, weights = farm_iptw$weights
)
```

When IPTW is used to estimate the causal treatment effect, a robust standard error estimate is required.

Estimate the regression parameters for the weighted regression model using the coeftest() function. Incorporate a robust standard error estimator from the sandwich package.

```{r}
# Estimate regression coefficients
# for weighted outcome model with
# robust standard errors
coeftest(
  yield_mod, vcov. = vcovHC
)
```

## Results Interpretation:

My estimate of the ATE is about 3.58. I estimate that when counties have at least 10% of farms with cover crops, average total yield of wheat increases by about 3.58 bushels per acre.

Conclusion

The results of this analysis suggest that the use of cover crops has a positive impact on wheat yields. Specifically, the estimated Average Treatment Effect (ATE) indicates that counties where at least 10% of farms adopt cover crops experience an increase in wheat yields of approximately 3.58 bushels per acre compared to counties with lower adoption rates. This finding highlights the potential benefits of cover crops not only for soil health but also for crop productivity.

However, it is important to note several limitations of this study. The analysis relies on observational data, which may still be subject to unmeasured confounding despite the use of causal inference methods. Additionally, the generalizability of these results may be limited by regional differences in farming practices and environmental conditions. Future research could benefit from more granular data and experimental studies to further validate these findings.

Overall, this project demonstrates the utility of causal inference techniques in agricultural research and provides evidence supporting the use of cover crops as a beneficial practice for enhancing wheat yields. These insights can inform both policymakers and farmers in their efforts to implement more sustainable and productive farming practices.